---
title: "Data visualization"
author: "Oleksandr Koval"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mission

This document created to make easier understanding data that we are working.

# Introduction

In this document we will observe some tables:

-   **ds_year** (describes the number of titles and the number of copies for each year, since 2018)

-   **ds_genre** (describes the number of titles and the number of copies for each year, by genre)

-   **ds_georaphy** (describes the number of titles and the number of copies for each year, by region)

-   **ds_language** (describes the number of titles and the number of copies for each year, by language)

-   **ds_pubtype** (describes the number of titles and the number of copies for each year, by publication type)

-   **ds_ukr_rus** (describes the number of titles and the number of copies, and the percentage of Ukrainian and Russian books for each year)

***P.s:*** All tables contain information for 2005–2024, except for *ds_language*

# Preparing

Load the libraries used in this report.

```{r}
#| label: preparing
#| code-summary: load libraries
#| echo: true
#| results: hide
#| message: false
#| code-fold: true

library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)

```

# Data import

We will be working with data that describes two main measures: the number of titles (**title_count**) and the number of copies (**copy_count**). Several specific tables provide these measures under different conditions.

```{r}
#| label: data-import
#| code-summary: Load data
#| echo: true
#| results: hide
#| code-fold: true

getwd()
ds_genre <- readRDS("../../data-private/derived/manipulation/ds_genre.rds") # load ds_genre

ds_year <- readRDS("../../data-private/derived/manipulation/ds_year.rds") # load ds_year

ds_language  <- readRDS("../../data-private/derived/manipulation/ds_language.rds") # load ds_language

ds_geography  <- readRDS("../../data-private/derived/manipulation/ds_geography.rds") # load ds_geography

ds_pubtype  <- readRDS("../../data-private/derived/manipulation/ds_pubtype.rds") # load ds_pubtype

ds_ukr_rus  <- readRDS("../../data-private/derived/manipulation/ds_ukr_rus.rds") # load ds_ukr_rus


```

# What we going to do?

We will create a short visualization for each table to make the data easier to understand. You will see several graphs that describe the main features of the data.

# Lets start !

## DS_YEAR

This table has `r nrow(ds_year)` rows and `r ncol(ds_year)` columns. In the **measure** column, there are only two possible values: **title_count** (the number of titles) and **copy_count** (the number of copies). The value column contains the numeric data for both measures.

The first graph shows the number of copies for each year.

```{r}
#| label: g_year_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 12
#| fig-height: 10


years_all <- seq(min(ds_year$year), max(ds_year$year))

g_year_copy <- ds_year %>% 
  filter(measure == "copy_count") %>% 
  ggplot(aes(x = year, y = value)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = years_all) +             
  scale_y_continuous(breaks = seq(0, 80000, by = 10000), limits = c(0, 80000)) +  
  labs(
    title = "Number of Copies by Year"
    , x = "The year"
    , y = "Number of copies (ths.)"
  )+
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
    ,panel.border = element_rect(color = "black", fill = NA, size = 1)
    ) 

g_year_copy_plotly <- ggplotly(g_year_copy)

g_year_copy_plotly

```

The second graph observe the number of titles for each year.

```{r}
#| label: g_year_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 12
#| fig-height: 10


g_year_title <- ds_year %>% 
  filter(
    measure == "title_count"
  ) %>% 
  ggplot(aes(x = year, y = value)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = years_all) +
  scale_y_continuous(breaks = seq(0, 27500, by = 5000), limits = c(7500, 27500)) +
  theme_minimal() +
  labs(
    title = "Number of Titles by Year"
    , x = "The year"
    , y = "Number of titles"
  ) +
  theme(panel.grid.minor = element_blank()
        ,panel.border = element_rect(color = "black", fill = NA, size = 1)) 


g_year_title_plotly <- ggplotly(g_year_title)

g_year_title_plotly

rm(g_year_title, g_year_copy, years_all)

```

What can we observe?

-   The most productive year for new titles and the year with the most copies

-   The overall production trend

## DS_GENRE

This table has `r nrow(ds_genre)` rows and `r ncol(ds_genre)` columns. In the **measure** column, there are only two possible values: **title_count** (the number of titles) and **copy_count** (the number of copies). Starting from the third column, each column represents a separate genre.

In the first graph, we see the number of copies for each genre, by year.

```{r}
#| label: g_genre_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 30



ds_genre_copy <- 
  ds_genre %>% 
  filter(
    measure == "copy_count"
  ) %>% 
  group_by(year)  %>%
  pivot_longer(
    cols = -c(year, measure),  # adding genre column
    names_to = "genre",
    values_to = "value"
  )


g_genre_copy<- ds_genre_copy %>%
  ggplot(aes(x = genre, y = value, fill = genre)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 30000, by = 7500), limits = c(0, 37000)) +
  labs(
    title = "Number of Copies by Genre and Year",
    x = "Genre",
    y = "Number of Copies (ths.)",
    fill = "Genre"
  ) +
  theme_get() +
  theme(
    axis.text.x = element_blank(),      
    axis.title.x = element_blank() 
    ,legend.position = "bottom"
    , panel.spacing.y = unit(5, "lines")
  )

g_genre_copy_plotly <- ggplotly(g_genre_copy) %>%
  layout(
    legend = list(
      orientation = "h",           
      x = 0.5,                     
      y = -0.1,                   
      xanchor = "center"
    )
  )

g_genre_copy_plotly
 
```

In a second graph, we can see the number of titles for each genre, by year.

```{r}
#| label: g_genre_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 30



ds_genre_title <- 
  ds_genre %>% 
  filter(
    measure == "title_count"
  ) %>% 
  group_by(year) %>%
  pivot_longer(
    cols = -c(year, measure),  # adding genre column
    names_to = "genre",
    values_to = "value"
  )

g_genre_title <- ds_genre_title %>%
  ggplot(aes(x = genre, y = value, fill = genre)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  labs(
    title = "Number of Titles by Genre and Year",
    x = "Genre",
    y = "Number of Titles",
    fill = "Genre"
  ) +
  theme_get() +
  theme(
    axis.text.x = element_blank(),      
    axis.title.x = element_blank()  
    ,legend.position = "bottom"
    ,  panel.spacing.y = unit(5, "lines"),

  )

g_genre_title_plotly <- ggplotly(g_genre_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_genre_title_plotly

rm(g_genre_copy, ds_genre_copy, ds_genre_title, g_genre_title)

```

## DS_GEOGRAPHY

This table has `r nrow(ds_geography)` rows and `r ncol(ds_geography)` columns. In the **measure** column, there are only two possible values: **title_count** (the number of titles) and **copy_count** (the number of copies). Starting from the third column, each column represents a separate geographical region.

```{r}
#| label: geography-preparing
#| code-summary: do some preparing
#| echo: true
#| results: hide
#| message: false
#| code-fold: true

region_groups <- list(
  "Західна Україна" = c("Львівська", "Івано-Франківська", "Закарпатська", "Тернопільська", "Чернівецька", "Волинська", "Рівненська"),
  "Центральна Україна" = c("Київська", "Житомирська", "Черкаська", "Кіровоградська", "Полтавська", "Вінницька"),
  "Південна Україна" = c("Одеська", "Миколаївська", "Херсонська", "Запорізька"),
  "Східна Україна" = c("Харківська", "Донецька", "Луганська", "Дніпропетровська"),
  "Північна Україна" = c("Чернігівська", "Сумська")
  , "Крим" = "Автономна Республіка Крим"
)

region_group_df <- tibble::tibble(
  region_name = unlist(region_groups),
  group = rep(names(region_groups), times = sapply(region_groups, length))
)

g_geography <- ds_geography %>%
  pivot_longer(
    cols = -c(year, measure),    
    names_to = "region_name",
    values_to = "value"
  )  %>%
  left_join(region_group_df, by = "region_name") %>%
  select(year, measure, region_name, group, everything())

```

## Inspecting the number of copies

### Central Ukraine (Copy count)

This graph describes the number of copies for each region in **Central Ukraine**

```{r}
#| label: g_central_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_central_copies <- g_geography %>%
  filter(measure == "copy_count", group == "Центральна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 750, by = 250)) +      
  coord_cartesian(ylim = c(0, 750)) +                       
  labs(
    title = "Number of Copies by Region (Central Ukraine) and Year",
    x = "Region",
    y = "Number of Copies (ths.)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
     ,  panel.spacing.y = unit(6, "lines")
     ,  panel.spacing.x = unit(2, "lines")

  )

g_geography_central_copies_plotly <- ggplotly(g_geography_central_copies) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_central_copies_plotly

rm(g_geography_central_copies)
```

### North Ukraine (Copy count)

This graph describes the number of copies for each region in **North Ukraine**

```{r}
#| label: g_north_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_pivnichna_copies <- g_geography %>%
  filter(measure == "copy_count", group == "Північна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 500, by = 250)) +      
  coord_cartesian(ylim = c(0, 500)) +                       
  labs(
    title = "Number of Copies by Region (North Ukraine) and Year",
    x = "Region",
    y = "Number of Copies (ths.)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_pivnichna_copies_plotly <- ggplotly(g_geography_pivnichna_copies) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_pivnichna_copies_plotly

rm(g_geography_pivnichna_copies)

```

### South Ukraine (Copy count)

This graph describes the number of copies for each region in **South Ukraine**

```{r}
#| label: g_south_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_pivdena_copies <- g_geography %>%
  filter(measure == "copy_count", group == "Південна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 500, by = 250)) +      
  coord_cartesian(ylim = c(0, 500)) +                       
  labs(
    title = "Number of Copies by Region (South Ukraine) and Year",
    x = "Region",
    y = "Number of Copies (ths.)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_pivdena_copies_plotly <- ggplotly(g_geography_pivdena_copies) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_pivdena_copies_plotly

rm(g_geography_pivdena_copies)

```

### West Ukraine (Copy count)

This graph describes the number of copies for each region in **West Ukraine**

```{r}
#| label: g_west_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25
g_geography_zahidna_copies <- g_geography %>%
  filter(measure == "copy_count", group == "Західна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 1250, by = 500)) +      
  coord_cartesian(ylim = c(0, 1250)) +                       
  labs(
    title = "Number of Copies by Region (West Ukraine) and Year",
    x = "Region",
    y = "Number of Copies (ths.)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_zahidna_copies_plotly <- ggplotly(g_geography_zahidna_copies) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_zahidna_copies_plotly

rm(g_geography_zahidna_copies)
```

### East Ukraine (Copy count)

This graph describes the number of copies for each region in **East Ukraine**

```{r}
#| label: g_east_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25
g_geography_shidna_copies <- g_geography %>%
  filter(measure == "copy_count", group == "Східна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 3500, by = 1000)) +      
  coord_cartesian(ylim = c(0, 3500)) +                       
  labs(
    title = "Number of Copies by Region (East Ukraine) and Year",
    x = "Region",
    y = "Number of Copies (ths.)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_shidna_copies_plotly <- ggplotly(g_geography_shidna_copies) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_shidna_copies_plotly

rm(g_geography_shidna_copies)
```

### Crimea (Copy count)

This graph describes the number of copies for **Crimea**

```{r}
#| label: g_krym_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 20
g_geography_krym_copies <- 
  g_geography %>% 
  filter(
    measure == "copy_count",
    group == "Крим"
  ) %>% 
  ggplot(aes(x = year, y = value, group = 1)) +   # add group=1 for geom_line
  geom_point() +
  geom_line() +
  labs(
    title = "Number of Copies by Region (Krym) and Year",
    x = "Year",
    y = "Number of Copies (ths.)"
  ) +
  scale_y_continuous(breaks = seq(0, 1500, by = 250), limits = c(0, 1500)) +  
  scale_x_continuous(breaks = 2005:2024) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"  # No legend needed, since only one line/region
  )

g_geography_krym_copies_plotly <- ggplotly(g_geography_krym_copies)

g_geography_krym_copies_plotly
rm(g_geography_krym_copies)
```

## Inspectin the number of titles

### Central Ukraine (Titles count)

This graph describes the number of titles for each region in **Central Ukraine**

```{r}
#| label: g_central_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_central_title <- g_geography %>%
  filter(measure == "title_count", group == "Центральна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 750, by = 250)) +      
  coord_cartesian(ylim = c(0, 750)) +                       
  labs(
    title = "Number of Titles by Region (Central Ukraine) and Year",
    x = "Region",
    y = "Number of Titles",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_central_title_plotly <- ggplotly(g_geography_central_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_central_title_plotly
rm(g_geography_central_title)
```

### North Ukraine (Titles count)

This graph describes the number of titles for each region in **North Ukraine**

```{r}
#| label: g_north_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_pivnichna_title <- g_geography %>%
  filter(measure == "title_count", group == "Північна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 750, by = 250)) +      
  coord_cartesian(ylim = c(0, 750)) +                       
  labs(
    title = "Number of Titles by Region (North Ukraine) and Year",
    x = "Region",
    y = "Number of Titles",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_pivnichna_title_plotly <- ggplotly(g_geography_pivnichna_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_pivnichna_title_plotly

rm(g_geography_pivnichna_title)
```

### South Ukraine (Titles count)

This graph describes the number of titles for each region in **South Ukraine**

```{r}
#| label: g_south_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_pivdena_title <- g_geography %>%
  filter(measure == "title_count", group == "Південна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 750, by = 250)) +      
  coord_cartesian(ylim = c(0, 750)) +                       
  labs(
    title = "Number of Titles by Region (South Ukraine) and Year",
    x = "Region",
    y = "Number of Titles",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_pivdena_title_plotly <- ggplotly(g_geography_pivdena_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_pivdena_title_plotly

rm(g_geography_pivdena_title)
```

### West Ukraine (Titles count)

This graph describes the number of titles for each region in **West Ukraine**

```{r}
#| label: g_west_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25

g_geography_zahidna_title <- g_geography %>%
  filter(measure == "title_count", group == "Західна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 1250, by = 500)) +      
  coord_cartesian(ylim = c(0, 1250)) +                       
  labs(
    title = "Number of Titles by Region (West Ukraine) and Year",
    x = "Region",
    y = "Number of Titles",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_zahidna_title_plotly <- ggplotly(g_geography_zahidna_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_zahidna_title_plotly

rm(g_geography_zahidna_title)
```

### East Ukraine (Titles count)

This graph describes the number of titles for each region in **East Ukraine**

```{r}
#| label: g_east_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 25
g_geography_shidna_title <- g_geography %>%
  filter(measure == "title_count", group == "Східна Україна") %>%
  complete(year, region_name, fill = list(value = 0)) %>%
  ggplot(aes(x = region_name, y = value, fill = region_name)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 4000, by = 1000)) +      
  coord_cartesian(ylim = c(0, 4000)) +                       
  labs(
    title = "Number of Titles by Region (East Ukraine) and Year",
    x = "Region",
    y = "Number of Titles",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.spacing.y = unit(6, "lines"),
    panel.spacing.x = unit(2, "lines")
  )

g_geography_shidna_title_plotly <- ggplotly(g_geography_shidna_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_geography_shidna_title_plotly

rm(g_geography_shidna_title)

```

### Crimea (Titles count)

This graph describes the number of titles for **Crimea**

```{r}
#| label: g_krym_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 20
g_geography_krym_title <- 
  g_geography %>% 
  filter(
    measure == "title_count",
    group == "Крим"
  ) %>% 
  ggplot(aes(x = year, y = value, group = 1)) +  # group=1 for line plot
  geom_point() +
  geom_line() +
  labs(
    title = "Number of Titles by Region (Krym) and Year",
    x = "Year",
    y = "Number of Titles"
  ) +
  scale_y_continuous(breaks = seq(0, 1000, by = 250), limits = c(0, 1000)) +  
  scale_x_continuous(breaks = 2005:2024) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "none"  # no legend needed
  )

g_geography_krym_title_plotly <- ggplotly(g_geography_krym_title)

g_geography_krym_title_plotly

rm(g_geography_krym_title)
```

## DS_LANGUAGE

This table has `r nrow(ds_language)` rows and `r ncol(ds_language)` columns. In the **measure** column, there are only two possible values: **title_count** (the number of titles) and **copy_count** (the number of copies). Starting from the third column, each column represents a new language. Data is available only for the years **2018–2024**.

To analyze this table, we will examine each year separately. Each graph displays only those languages that have at least 10 different titles and 10,000 copies.

```{r}
#| label: language-preparing
#| code-summary: do some preparing
#| echo: true
#| results: hide
#| message: false
#| code-fold: true

g_language <- ds_language %>%
  pivot_longer(
    cols = -c(year, measure),     
    names_to = "language",
    values_to = "value"
  )

```

### 2018 (Language)

```{r}
#| label: g_language_2018
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2018 <- 
  g_language %>% 
  filter(
    year == 2018,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька") 
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 40000, by = 5000), limits = c(0, 40000)) +  
  labs(
    title = "Number of Copies and Titles (2018)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
    , axis.text.x = element_text(size = 8) 
  )

g_language_2018_plotly <- ggplotly(g_language_2018) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2018_plotly

rm(g_language_2018)

```

### 2019 (Language)

```{r}
#| label: g_language_2019
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2019 <- 
  g_language %>% 
  filter(
    year == 2019,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    # Rename languages as needed
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька")
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 55000, by = 10000), limits = c(0, 55000)) +  
  labs(
    title = "Number of Copies and Titles (2019)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8)
  )

g_language_2019_plotly <- ggplotly(g_language_2019) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2019_plotly

rm(g_language_2019)

```

### 2020 (Language)

```{r}
#| label: g_language_2020
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2020 <- 
  g_language %>% 
  filter(
    year == 2020,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    # Rename language as needed:
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька")
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 40000, by = 5000), limits = c(0, 40000)) +  
  labs(
    title = "Number of Copies and Titles (2020)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8)
  )

g_language_2020_plotly <- ggplotly(g_language_2020) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2020_plotly

rm(g_language_2020)

```

### 2021 (Language)

```{r}
#| label: g_language_2021
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2021 <- 
  g_language %>% 
  filter(
    year == 2021,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    # Rename language as needed:
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька")
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 40000, by = 5000), limits = c(0, 40000)) +  
  labs(
    title = "Number of Copies and Titles (2021)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8)
  )

g_language_2021_plotly <- ggplotly(g_language_2021) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2021_plotly

rm(g_language_2021)
```

### 2022 (Language)

```{r}
#| label: g_language_2022
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2022 <- 
  g_language %>% 
  filter(
    year == 2022,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    # Optional: rename any language here
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька")
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 15000, by = 5000), limits = c(0, 15000)) +  
  labs(
    title = "Number of Copies and Titles (2022)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8)
  )

g_language_2022_plotly <- ggplotly(g_language_2022) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2022_plotly


rm(g_language_2022)
```

### 2023 (Language)

```{r}
#| label: g_language_2023
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2023 <- 
  g_language %>% 
  filter(
    year == 2023,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    # Optional: rename any language as needed
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька")
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 25000, by = 5000), limits = c(0, 25000)) +  
  labs(
    title = "Number of Copies and Titles (2023)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8)
  )

g_language_2023_plotly <- ggplotly(g_language_2023) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2023_plotly

rm(g_language_2023)
```

### 2024 (Language)

```{r}
#| label: g_language_2024
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 14

g_language_2024 <- 
  g_language %>% 
  filter(
    year == 2024,
    (measure == "copy_count" & value > 10) | (measure == "title_count" & value > 10)
  ) %>%
  mutate(
    measure = recode(measure,
                     copy_count = "Copies",         
                     title_count = "Titles"),
    # Optional: rename any language as needed
    language = recode(language, 
                      `Кількома мовами народів світу` = "Декілька")
  ) %>%
  ggplot(aes(x = language, y = value, fill = measure)) +
  geom_col(position = "dodge") +
  scale_y_continuous(breaks = seq(0, 35000, by = 5000), limits = c(0, 35000)) +  
  labs(
    title = "Number of Copies and Titles (2024)",
    x = "Language",
    y = "Count",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 8)
  )

g_language_2024_plotly <- ggplotly(g_language_2024) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_language_2024_plotly

rm(g_language_2024)

```

## DS_PUBTYPE

This table has `r nrow(ds_pubtype)` rows and `r ncol(ds_pubtype)` columns. In the **measure** column, there are only two possible values: *title_count* (the number of titles) and *copy_count* (the number of copies). Starting from the third column, each column represents a different publication type.

The first graph shows the number of copies for each publication type, by year.

```{r}
#| label: g_pubtype_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 30


ds_pubtype_l <- ds_pubtype %>%
  pivot_longer(
    cols = -c(year, measure),  
    names_to = "pubtype",      
    values_to = "value"       
  )


ds_pubtype_copy <- 
  ds_pubtype_l %>%
  filter(measure == "copy_count") %>%
  group_by(year)

g_pubtype_copy <- ds_pubtype_copy %>%
  ggplot(aes(x = pubtype, y = value, fill = pubtype)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 30000, by = 7500), limits = c(0, 37000)) +
  coord_cartesian(ylim = c(0, 37000)) +
  labs(
    title = "Number of Copies by Publication Type and Year",
    x = "Publication Type",
    y = "Number of Copies (ths.)",
    fill = "Publication Type"
  ) +
  theme_get() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
        ,legend.position = "bottom"
    , panel.spacing.y = unit(5, "lines")
  )

g_pubtype_copy_plotly <- ggplotly(g_pubtype_copy) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_pubtype_copy_plotly

rm(ds_pubtype_l, ds_pubtype_copy, g_pubtype_copy)

```

Second graph describes the number of titles of each publication type, by year.

```{r}
#| label: g_pubtype_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 20
#| fig-height: 30

ds_pubtype_l <- ds_pubtype %>%
  pivot_longer(
    cols = -c(year, measure),  
    names_to = "pubtype",      
    values_to = "value"       
  )

ds_pubtype_title <- 
  ds_pubtype_l %>%
  filter(measure == "title_count") %>%
  group_by(year)

g_pubtype_title <- ds_pubtype_title %>%
  ggplot(aes(x = pubtype, y = value, fill = pubtype)) +
  geom_col() +
  facet_wrap(~ year, ncol = 4) +
  scale_y_continuous(breaks = seq(0, 9000, by = 2500), limits = c(0, 9000)) +
  coord_cartesian(ylim = c(0, 9000)) +
  labs(
    title = "Number of Titles by Publication Type and Year",
    x = "Publication Type",
    y = "Number of Titles",
    fill = "Publication Type"
  ) +
  theme_get() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "bottom",
    panel.spacing.y = unit(5, "lines")
  )

g_pubtype_title_plotly <- ggplotly(g_pubtype_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.1,
      xanchor = "center"
    )
  )

g_pubtype_title_plotly

rm(ds_pubtype_title, g_pubtype_title)


```

## DS_UKR_RUS

This table has `r nrow(ds_pubtype)` rows and `r ncol(ds_pubtype)` columns. In the **measure** column, there are only two possible values: *title_count* (the number of titles) and *copy_count* (the number of copies). The third column shows the percentage of Ukrainian books among Ukrainian and Russian editions. The fourth column shows the percentage of Russian books among Ukrainian and Russian editions.

The first graph shows the number of copies of Ukrainian and Russian books for each year from 2005 to 2024.

```{r}
#| label: g_ukrrus_copy
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 16


g_ukrrus_copies <- 
  ds_ukr_rus %>%
  filter(measure == "copy_count") %>%
  select(year, ukr, rus) %>%
  pivot_longer(
    cols = c(ukr, rus),
    names_to = "language",
    values_to = "value"
  ) %>% 
  ggplot(aes(x = factor(year), y = value, fill = language)) +
  geom_col(position = "dodge") +
  scale_x_discrete(expand = expansion(mult = c(0))) +
  labs(
    title = "Number of Copies: Ukrainian vs Russian by Year",
    x = "Year",
    y = "Number of Copies (ths.)",
    fill = "Language"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

g_ukrrus_copies_plotly <- ggplotly(g_ukrrus_copies) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_ukrrus_copies_plotly


rm(g_ukrrus_copies)
```

Second graph describes the number of titles of Ukrainian and russian titles for each year since 2005 to 2024.

```{r}
#| label: g_ukrrus_title
#| code-summary: Making graph
#| results: markup
#| message: false
#| code-fold: true
#| warning: false
#| fig-width: 15
#| fig-height: 16


g_ukrrus_title <- 
  ds_ukr_rus %>%
  filter(measure == "title_count") %>%
  select(year, ukr, rus) %>%
  pivot_longer(
    cols = c(ukr, rus),
    names_to = "language",
    values_to = "value"
  ) %>% 
  ggplot(aes(x = factor(year), y = value, fill = language)) +
  geom_col(position = "dodge") +
  scale_x_discrete(expand = expansion(mult = c(0))) +
  scale_y_continuous(breaks = seq(0, 20000, by = 5000), limits = c(0, 20000)) +
  labs(
    title = "Number of Titles: Ukrainian vs Russian by Year",
    x = "Year",
    y = "Number of Titles",
    fill = "Language"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

g_ukrrus_title_plotly <- ggplotly(g_ukrrus_title) %>%
  layout(
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.15,
      xanchor = "center"
    )
  )

g_ukrrus_title_plotly

rm(g_ukrrus_title)
```

# Overral

This document was created to help you gain a deeper understanding of the data. I hope you found it useful! 😉