---
title: "Exploratory Data Analysis: Books of Ukraine"
subtitle: "Publishing Trends and Language Dynamics 2005-2023"
author: "Research & Open Data"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mission

This document explores publishing trends in Ukraine from 2005 to 2023, with particular attention to:

-   **Temporal patterns** in book publishing volume and circulation
-   **Language dynamics** between Ukrainian and Russian publications
-   **Geographic distribution** of publishing activities across Ukrainian regions
-   **Genre and publication type** patterns over time

The analysis focuses on understanding regional differences and detecting interesting patterns and relationships between Russian language use in published books and the larger cultural, political, and economic context of Ukraine.

# **Definition of Terms**

Following the project ontology, this analysis works with several key measurement dimensions:

-   **Publication Count** (`titles_count`) - Number of unique book titles published
-   **Copy Count** (`copy_count`) - Total print run/circulation of publications
-   **Geographic Regions** - Ukrainian oblasts grouped into macro-regions (North, South, East, West, Center)
-   **Language Categories** - Ukrainian, Russian, English, and other languages
-   **Genre Structure** - 13 thematic categories from political literature to children's books
-   **Publication Types** - 14 categories from scientific to religious publications

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false 
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir='../../') # keep this chunk isolated  
```

```{r}
#| label: set_options
#| code-summary: report-wide formatting options
#| echo: false
#| results: hide
#| message: false 
cat("Working directory: ", getwd()) # turn on to test the location
report_render_start_time <- Sys.time()
# set options shared by all chunks
opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  # dpi        = 400, # dots per inch,
  out.width  = "960px", # pixels, this affects only the markdown, not the underlying png file.  The height will be scaled appropriately.
  fig.width    = 10, # inches
  fig.height   = 6, # inches
  fig.path     = 'figure-png-iso/' # where figures are stored
  
)
echo_chunks    <- FALSE #Toggle for debugging.
message_chunks <- FALSE #Toggle for debugging.
options(width=100) # number of characters to display in the output (dflt = 80)
ggplot2::theme_set(ggplot2::theme_bw()) # common theme for all graphs
read_chunk("./analysis/eda-1/eda-1.R") #This allows knitr to call
```

# Environment

> Reviews the components of the working environment of the report. Non-technical readers are welcomed to skip. Come back if you need to understand the origins of custom functions, scripts, or data objects.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true

```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: define-query
#| code-summary: Define queries to run
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

This report operates with the following data objects:

-   **`ds_year`** - Annual summary data for total publications and copies
-   **`ds_language`** - Publication counts by language (Ukrainian, Russian, English, etc.)\
-   **`ds_genre`** - Distribution across 13 genre categories
-   **`ds_pubtype`** - Distribution across 14 publication types
-   **`ds_geography`** - Regional distribution across Ukrainian oblasts
-   **`ds_ukr_rus`** - Focused comparison between Ukrainian and Russian publications


```{r }
#| label: load-data
#| code-summary: Load data objects from manipulation pipeline
#| warning: false
#| code-fold: true
#| cache: true
```

## Data Availability Check

```{r}
#| label: inspect-data-0
#| code-summary: Verify all required datasets are available
#| code-fold: false
#| eval: true
#| cache: true

# Display basic information about loaded datasets
cat("Dataset dimensions:\n")
cat("ds_year:", dim(ds_year)[1], "rows x", dim(ds_year)[2], "columns\n")
cat("ds_language:", dim(ds_language)[1], "rows x", dim(ds_language)[2], "columns\n") 
cat("ds_genre:", dim(ds_genre)[1], "rows x", dim(ds_genre)[2], "columns\n")
cat("ds_pubtype:", dim(ds_pubtype)[1], "rows x", dim(ds_pubtype)[2], "columns\n")
cat("ds_geography:", dim(ds_geography)[1], "rows x", dim(ds_geography)[2], "columns\n")
cat("ds_ukr_rus:", dim(ds_ukr_rus)[1], "rows x", dim(ds_ukr_rus)[2], "columns\n")

# Show year range
cat("\nYear range in data:", min(ds_year$yr), "-", max(ds_year$yr), "\n")

# Show measure types
cat("\nMeasure types available:", paste(unique(ds_year$measure), collapse = ", "), "\n")
```

## Data Structure Overview

```{r}
#| label: tweak-data-0
#| code-summary: Prepare enhanced datasets for analysis
#| code-fold: true
#| cache: true
#| results: hide
```

### Annual Totals Structure

```{r}
#| code-summary: Overview of annual summary data
#| code-fold: false
ds_year %>% 
  head(10) %>%
  knitr::kable(caption = "Sample of Annual Summary Data")
```

### Language Data Structure

```{r}
#| code-summary: Overview of language distribution data
#| code-fold: false
ds_language %>% 
  select(1:6) %>% # Show first few columns
  head(6) %>%
  knitr::kable(caption = "Sample of Language Distribution Data")
```

### Ukrainian-Russian Focus Data

```{r}
#| code-summary: Overview of Ukrainian-Russian comparison data
#| code-fold: false  
ds_ukr_rus %>%
  head(8) %>%
  knitr::kable(caption = "Ukrainian vs Russian Publications Data")
```

# Ready for Analysis

The data is now loaded and prepared for human analysts to:

1.  **Create temporal visualizations** showing publishing trends over time
2.  **Analyze language dynamics** between Ukrainian and Russian publications\
3.  **Explore geographic patterns** across Ukrainian regions
4.  **Investigate genre and publication type trends**
5.  **Correlate patterns** with historical events and policy changes

## Available Helper Functions

-   `create_time_series()` - Generate time series plots for any numeric variable
-   `create_language_comparison()` - Compare Ukrainian vs Russian publication patterns
-   `connect_to_db()` - Access SQLite database for custom queries

## Next Steps for Human Analysts

```{r}
#| code-summary: Template for creating your first visualization
#| eval: false
#| code-fold: false

# Example 1: Overall publishing trends
ds_year %>%
  filter(measure == "title_count") %>%
  create_time_series("value", 
                    title = "Total Book Publications in Ukraine",
                    subtitle = "Annual count of new titles published",
                    y_label = "Number of Titles")

# Example 2: Language comparison over time  
create_language_comparison(ds_language, "title_count")

# Example 3: Custom analysis
# Add your own code here for specific research questions
```

------------------------------------------------------------------------

**Report generated:** `r Sys.time()`\
**Analysis period:** `r target_window_opens` to `r target_window_closes`\
**Data source:** Ukrainian publishing statistics via manipulation/0-ellis.R pipeline
