---
title: "Exploratory Data Analysis: Books of Ukraine"
subtitle: "Publishing Trends and Language Dynamics 2005-2023"
author: "Research & Open Data"
format:
  html:
    page-layout: full
    toc: true
    toc-location: right
    code-fold: show
    theme: yeti
    highlight-style: breeze
    code-line-numbers: true
    self-contained: true
    embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
---

# Mission

This document explores publishing trends in Ukraine from 2005 to 2023, with particular attention to:

-   **Temporal patterns** in book publishing volume and circulation
-   **Language dynamics** between Ukrainian and Russian publications
-   **Geographic distribution** of publishing activities across Ukrainian regions
-   **Genre and publication type** patterns over time

The analysis focuses on understanding regional differences and detecting interesting patterns and relationships between Russian language use in published books and the larger cultural, political, and economic context of Ukraine.

# **Definition of Terms**

Following the project ontology, this analysis works with several key measurement dimensions:

-   **Publication Count** (`titles_count`) - Number of unique book titles published
-   **Copy Count** (`copy_count`) - Total print run/circulation of publications
-   **Geographic Regions** - Ukrainian oblasts grouped into macro-regions (North, South, East, West, Center)
-   **Language Categories** - Ukrainian, Russian, English, and other languages
-   **Genre Structure** - 13 thematic categories from political literature to children's books
-   **Publication Types** - 14 categories from scientific to religious publications

```{r}
#| code-summary: Reconciles working directory between execution modes
#| echo: false
#| results: hide
#| message: false 
#| warning: false
cat("Working directory: ", getwd())
library(knitr)
opts_knit$set(root.dir='../../') # keep this chunk isolated  
```

```{r}
#| label: set_options
#| code-summary: report-wide formatting options
#| echo: false
#| results: hide
#| message: false 
cat("Working directory: ", getwd()) # turn on to test the location
report_render_start_time <- Sys.time()
# set options shared by all chunks
opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  # dpi        = 400, # dots per inch,
  out.width  = "960px", # pixels, this affects only the markdown, not the underlying png file.  The height will be scaled appropriately.
  fig.width    = 10, # inches
  fig.height   = 6, # inches
  fig.path     = 'figure-png-iso/' # where figures are stored
  
)
echo_chunks    <- FALSE #Toggle for debugging.
message_chunks <- FALSE #Toggle for debugging.
options(width=100) # number of characters to display in the output (dflt = 80)
ggplot2::theme_set(ggplot2::theme_bw()) # common theme for all graphs
read_chunk("./analysis/eda-1/eda-1.R") #This allows knitr to call
```

# Environment

> Reviews the components of the working environment of the report. Non-technical readers are welcomed to skip. Come back if you need to understand the origins of custom functions, scripts, or data objects.

```{r}
#| label: load-packages
#| code-summary: Packages used in the current report
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: load-sources
#| code-summary: Collection of custom functions used in current repository
#| echo: true
#| results: hide
#| message: false
#| code-fold: true

```

```{r}
#| label: declare-globals
#| code-summary: Values used throughout the report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: declare-functions
#| code-summary: Custom functions defined for use in this report.
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

```{r}
#| label: define-query
#| code-summary: Define queries to run
#| echo: true
#| results: hide
#| message: false
#| code-fold: true
```

# Data

This report operates with the following data objects in **long format**:

-   **`ds_year_long`** - Annual summary data in long format with measures for total publications and copies
-   **`ds_language_long`** - Publication counts by language in long format (Ukrainian, Russian, English, etc.)\
-   **`ds_genre_long`** - Distribution across 13 genre categories in long format
-   **`ds_pubtype_long`** - Distribution across 14 publication types in long format
-   **`ds_geography_long`** - Regional distribution across Ukrainian oblasts in long format
-   **`ds_ukr_rus_long`** - Focused comparison between Ukrainian and Russian publications in long format

All datasets follow the consistent long format structure with columns: `yr` (year), `measure` (metric type), `value` (count), and category-specific columns (e.g., `language`, `geography`, `genre`).


```{r }
#| label: load-data
#| code-summary: Load data objects from manipulation pipeline
#| warning: false
#| code-fold: true
#| cache: true
```

## Data Availability Check

```{r}
#| label: inspect-data-0
#| code-summary: Verify all required datasets are available
#| code-fold: false
#| eval: true
#| cache: true
#| echo: false

# Display basic information about loaded datasets (long format)
cat("Dataset dimensions (long format):\n")
cat("ds_year_long:", dim(ds_year_long)[1], "rows x", dim(ds_year_long)[2], "columns\n")
cat("ds_language_long:", dim(ds_language_long)[1], "rows x", dim(ds_language_long)[2], "columns\n") 
cat("ds_genre_long:", dim(ds_genre_long)[1], "rows x", dim(ds_genre_long)[2], "columns\n")
cat("ds_pubtype_long:", dim(ds_pubtype_long)[1], "rows x", dim(ds_pubtype_long)[2], "columns\n")
cat("ds_geography_long:", dim(ds_geography_long)[1], "rows x", dim(ds_geography_long)[2], "columns\n")
cat("ds_ukr_rus_long:", dim(ds_ukr_rus_long)[1], "rows x", dim(ds_ukr_rus_long)[2], "columns\n")

# Show year range
cat("\nYear range in data:", min(ds_year_long$yr), "-", max(ds_year_long$yr), "\n")

# Show measure types
cat("\nMeasure types available:", paste(unique(ds_year_long$measure), collapse = ", "), "\n")
```

## Data Structure Overview

```{r}
#| label: tweak-data-0
#| code-summary: Prepare enhanced datasets for analysis
#| code-fold: true
#| cache: true
#| results: hide
```

### Annual Totals Structure (Long Format)

```{r}
#| code-summary: Overview of annual summary data in long format
#| code-fold: false
ds_year_long %>% 
  head(10) %>%
  knitr::kable(caption = "Sample of Annual Summary Data (Long Format)")
```

### Language Data Structure (Long Format)

```{r}
#| code-summary: Overview of language distribution data in long format
#| code-fold: false
ds_language_long %>% 
  head(10) %>%
  knitr::kable(caption = "Sample of Language Distribution Data (Long Format)")
```

### Ukrainian-Russian Focus Data (Long Format)

```{r}
#| code-summary: Overview of Ukrainian-Russian comparison data in long format
#| code-fold: false  
ds_ukr_rus_long %>%
  head(10) %>%
  knitr::kable(caption = "Ukrainian vs Russian Publications Data (Long Format)")
```

# Ready for Analysis

The data is now loaded in **long format** and prepared for human analysts to:

1.  **Create temporal visualizations** showing publishing trends over time using consistent long format structure
2.  **Analyze language dynamics** between Ukrainian and Russian publications with `ds_language_long`\
3.  **Explore geographic patterns** across Ukrainian regions using `ds_geography_enhanced` with regional groupings
4.  **Investigate genre and publication type trends** using `ds_genre_long` and `ds_pubtype_long`
5.  **Correlate patterns** with historical events and policy changes across all dimensions

## Available Helper Functions (Long Format Optimized)

-   `create_time_series()` - Generate time series plots for any numeric variable from long format data
-   `create_language_comparison()` - Compare Ukrainian vs Russian publication patterns using `ds_language_long`
-   `create_regional_comparison()` - Compare trends across Ukrainian regions using `ds_geography_enhanced`
-   `analyze_category_trends()` - Flexible function to analyze any categorical variable from long format data
-   `connect_to_db()` - Access SQLite database for custom queries

## Advantages of Long Format

The long format structure provides several analytical advantages:

- **Consistent filtering**: All datasets use `measure` column to distinguish between `title_count` and `copy_count`
- **Easy aggregation**: Group by categorical variables and sum `value` for custom analysis
- **Flexible visualization**: Compatible with ggplot2 aesthetic mappings for complex multi-dimensional plots
- **Scalable analysis**: Same analytical patterns work across all categorical dimensions

## Next Steps for Human Analysts

```{r}
#| code-summary: Template for creating your first visualization with long format data
#| eval: false
#| code-fold: false

# Example 1: Overall publishing trends using long format
ds_year_long %>%
  filter(measure == "title_count") %>%
  create_time_series("value", 
                    title = "Total Book Publications in Ukraine",
                    subtitle = "Annual count of new titles published",
                    y_label = "Number of Titles")

# Example 2: Language comparison over time using long format 
create_language_comparison(ds_language_long, "title_count")

# Example 3: Regional analysis using long format
ds_geography_enhanced %>%
  filter(measure == "title_count") %>%
  group_by(yr, region_group) %>%
  summarise(total_titles = sum(value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = yr, y = total_titles, color = region_group)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Regional Publishing Trends", 
       x = "Рік", y = "Number of Titles", color = "Region") +
  theme_minimal()

# Example 4: Custom analysis with long format data
# Add your own code here for specific research questions
```

## Sample Analysis Chunks

The following chunks demonstrate the analytical power of long format data:

```{r}
#| label: sample-analysis
#| code-summary: Sample analyses demonstrating long format advantages
#| eval: false
#| code-fold: false
```

------------------------------------------------------------------------

**Report generated:** `r Sys.time()`\
**Analysis period:** `r target_window_opens` to `r target_window_closes`\
**Data source:** Ukrainian publishing statistics via manipulation/0-ellis.R pipeline
